---
title: "CSDE 502: Assignment #9"
author: "[Esteban J. Valencia](mailto:estebanv@uw.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d %H:%M")`'
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
output:
  bookdown::html_document2:
    theme: journal
    number_sections: TRUE
    self_contained: TRUE
    code_folding: hide
    toc: TRUE
    toc_float:
      collapsed: TRUE
      smooth_scroll: TRUE
urlcolor: blue

output: 
    bookdown::html_document2:
        theme: journal
        number_sections: true
        self_contained: true
        code_folding: hide
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
urlcolor: blue        
---
```{css, echo=FALSE}
#Step 0.0.1: Setting wide-wise scrolling
pre, code {white-space: pre !important; overflow-x: auto}

```

```{r setup, warning = FALSE, message = FALSE, results = "hide"}
#Step 0.0.2: Specifying chunk options
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      results = "hide",
                      width = 1000)

#Step 0.0.3: Loading requisite packages from library
library(captioner)
library(dplyr)
library(magrittr)
library(knitr)
library(kableExtra)

#Step 0.0.4: Setting captions
fig.nums = captioner(prefix = "Figure")
tab.nums = captioner(prefix = "Table")
```

# Overview

This assignment is intended to give you more practice delving into the Add Health data set and in manipulating additional variables.

Using the full household roster (you'll need to go back to teh full raw data source) create the following variables for each respondents: (1) total number in household, (2) number of sisters, (3) number of brothers, (4) and total number of siblings.

Additionally, calculate the proportion of students living with two biological parents. 

Additionally, calculate the number of household members that are *NOT* the biological mother, biological father, full brother, or full sister.

# Source Code Appendix

The source code for this document is included below.
```{r comment = ''}
cat(readLines(con = "estebanv_csde502_hw9.Rmd"), sep = "\n")
```