---
title: "CSDE 502: Assignment #9"
author: "[Esteban J. Valencia](mailto:estebanv@uw.edu)"
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
output:
  bookdown::html_document2:
    theme: journal
    number_sections: TRUE
    self_contained: TRUE
    code_folding: hide
    toc: TRUE
    toc_float:
      collapsed: TRUE
      smooth_scroll: TRUE
urlcolor: blue
---
```{css, echo=FALSE}
#Step 0.0.1: Setting wide-wise scrolling
pre, code {white-space: pre !important; overflow-x: auto}

```

```{r setup, warning = FALSE, message = FALSE, results = "hide"}
#Step 0.0.2: Specifying chunk options
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      results = "hide",
                      width = 1000)

#Step 0.0.4: Setting additional options
options(repos = "https://mirror.las.iastate.edu/CRAN/",
        width = 1000)

#Step 0.0.5: Loading requisite packages from library
library(captioner)
library(dplyr)
library(tidyverse)
library(magrittr)
library(knitr)
library(kableExtra)

#Step 0.0.6: Setting captions
fig.nums = captioner(prefix = "Figure")
tab.nums = captioner(prefix = "Table")
```

# Overview

This assignment is intended to give you more practice delving into the Add Health data set and in manipulating additional variables.

Using the full household roster (you'll need to go back to teh full raw data source) create the following variables for each respondents: (1) total number in household, (2) number of sisters, (3) number of brothers, (4) and total number of siblings.

Additionally, calculate the proportion of students living with two biological parents. 

Additionally, calculate the number of household members that are *NOT* the biological mother, biological father, full brother, or full sister.

## Loading Data

The requisite Add Health data are downloaded to a temporary directory and then unzipped, after which the temporary directory is closed and
```{r}
#Step 1.1.1: Downloading & opening Add Health zip-file from URL.

  ##Specifying URL
  f.url = "http://staff.washington.edu/phurvitz/csde502_winter_2021/data/21600-0001-Data.dta.zip"

  ##Setting temporary directory
  t.dir = Sys.getenv("TEMP")

  ##Setting path for zipfile
  z.file = file.path(t.dir, basename = "21600-0001-Data.dta.zip")
  
  ##Setting file in temporary folder
  f.data = tools::file_path_sans_ext(z.file)
  
  ##Check loop: does file exist?
  if(!file.exists(f.data)){
    
    ###If the file does not exist...
    if(!file.exists(z.file)){
      curl::curl_download(url = f.url, destfile = z.file)
    }
    
    ###If the file does exist...
    unzip(zipfile = z.file, exdir = t.dir)
  }
  
  ##Check loop: has the file been read? If not, read file...
  if(!exists("d.adh")) {
    d.adh = haven::read_dta(f.data)
  }
  
  ##Changing column names to lower case
  colnames(d.adh) %<>% str_to_lower()
```

# Total Number of Household Members

The Add Health data do not contain a variable indicating the total number of siblings in a household members. Instead, they list a series of variables detailing the relationship of each household member to the respondent. As such, to identify the total number of household members, we must parse through these variables. First, we subset the data to contain only those variables of relevance (in the Add Health data, those variables all begin with the prefix "H1HR").

```{r}
#Step 2.0.1: Subsetting variables
s.df = d.adh %>% select(starts_with("h1hr"))
```

Next, a proxy data frame is generated from the series of variables which ask about HH members' gender, and the total number of HH members is counted based on responses.

```{r}
#Step 2.0.2: Generating proxy data frame
p.df = s.df %>% select(starts_with("h1hr2"))

#Step 2.0.3: Recoding values based on codebook
p.df[p.df == 1 | p.df == 2 | p.df == 3 | p.df == 6 | p.df == 8] = 1
p.df[p.df == 7] = 0

#Step 2.0.4: Counting total number of responses indicating n'th HH member is present
p.df %<>% mutate(hh_mtotal = rowSums(p.df) + 1)

#Step 2.0.5: Binding results to main data frame
s.df %<>% mutate(hh_mtotal = p.df$hh_mtotal)

```

A tabulation of the total number of HH members in each HH is furnished below.

```{r}
#Step 2.0.6: Generating frequency tabulation

  ##Building table
  tab.x = data.frame(table(s.df$hh_mtotal))
  
  ##Renaming table columns
  names(tab.x) = c("Total number of household members", "Frequency")
  
  ##Formatting table
  tab.z = kable(tab.x) %>% 
          kable_styling(full_width = FALSE, 
                        position = "left", 
                        bootstrap_options = c("striped", "hover"))
```

`r tab.nums(name = "t01", caption = "Frequency Table, Total Number of Household Members, Including Respondent")`
```{r, echo = FALSE, results = "markdown"}
#Step 2.0.7: Printing table
tab.z
```

A histogram describing the total number of HH members in each HH is furnished below.

```{r}
#Step 2.0.8: Generating histogram
fig.z = ggplot(data = s.df, aes(hh_mtotal)) +
        geom_histogram(binwidth = 1) +
        scale_x_continuous(breaks = seq(1, 25, 1)) +
        xlab("\nTotal Number of Household Members, Including Respondent") + 
        ylab("Frequency Count\n")
```

`r fig.nums(name = "f01", caption = "Histogram, Total number of Household Members, Including Respondant")`
```{r, echo = FALSE, results = "markdown"}
#Step 2.0.9: Printing figure
fig.z
```

# Total number of Sisters

Similar to the process of counting the total number of household members, we create a proxy data set counting the relationship of each household member to the respondent, and then sum the total responses of "sister"

```{r}
#Step 3.0.1: Generating proxy data frame
p.df = s.df %>% select(starts_with("h1hr3"))

#Step 3.0.2: Recoding values based on codebook
p.df[p.df != 8] = 0
p.df[p.df == 8] = 1

#Step 3.0.3: Counting total number of responses indicating n'th HH member is present
p.df %<>% mutate(hh_stotal = rowSums(p.df))

#Step 3.0.4: Binding results to main data frame
s.df %<>% mutate(hh_stotal = p.df$hh_stotal)

#Step 3.0.5: Tabulating frequency

  ##Building table
  tab.x = data.frame(table(s.df$hh_stotal))
  
  ##Renaming table columns
  names(tab.x) = c("Total number of sisters", "Frequency")
  
  ##Formatting table
  tab.z = kable(tab.x) %>% 
          kable_styling(full_width = FALSE, 
                        position = "left", 
                        bootstrap_options = c("striped", "hover"))

#Step 3.0.6: Generating histogram
fig.z = ggplot(data = s.df, aes(hh_stotal)) +
        geom_histogram(binwidth = 1) +
        scale_x_continuous(breaks = seq(1, 25, 1)) +
        xlab("\nTotal Number of Sisters in Household") + 
        ylab("Frequency Count\n")
```

A frequency tabulation of the total number of sisters in the household is furnished below.

`r tab.nums(name = "t02", caption = "Frequency Table, Total Number of Sisters in Household")`
```{r, echo = FALSE, results = "markdown"}
#Step 3.0.7: Printing table
tab.z
```

A histogram describing the total number of sisters in the household is furnished below.
`r fig.nums(name = "f02", caption = "Histogram, Total Number of Sisters in Household")`
```{r, echo = FALSE, results = "markdown"}
#Step 3.0.7: Printing table
fig.z
```

# Total number of Brothers

Similar to the process of counting the total number of sisters, we create a proxy data set counting the relationship of each household member to the respondent, and then sum the total responses of "brother"

```{r}
#Step 4.0.1: Generating proxy data frame
p.df = s.df %>% select(starts_with("h1hr3"))

#Step 4.0.2: Recoding values based on codebook
p.df[p.df != 5] = 0
p.df[p.df == 5] = 1

#Step 4.0.3: Counting total number of responses indicating n'th HH member is present
p.df %<>% mutate(hh_btotal = rowSums(p.df))

#Step 4.0.4: Binding results to main data frame
s.df %<>% mutate(hh_btotal = p.df$hh_btotal)

#Step 4.0.5: Tabulating frequency

  ##Building table
  tab.x = data.frame(table(s.df$hh_btotal))
  
  ##Renaming table columns
  names(tab.x) = c("Total number of brothers", "Frequency")
  
  ##Formatting table
  tab.z = kable(tab.x) %>% 
          kable_styling(full_width = FALSE, 
                        position = "left", 
                        bootstrap_options = c("striped", "hover"))

#Step 4.0.6: Generating histogram
fig.z = ggplot(data = s.df, aes(hh_btotal)) +
        geom_histogram(binwidth = 1) +
        scale_x_continuous(breaks = seq(1, 25, 1)) +
        xlab("\nTotal Number of brothers in Household") + 
        ylab("Frequency Count\n")
```

A frequency tabulation of the total number of sisters in the household is furnished below.

`r tab.nums(name = "t03", caption = "Frequency Table, Total Number of Brothers in Household")`
```{r, echo = FALSE, results = "markdown"}
#Step 4.0.7: Printing table
tab.z
```

A histogram describing the total number of sisters in the household is furnished below.
`r fig.nums(name = "f03", caption = "Histogram, Total Number of Brothers in Household")`
```{r, echo = FALSE, results = "markdown"}
#Step 4.0.7: Printing table
fig.z
```

# Total number of siblings

To calculate the total number of siblings, we sum the total number of sisters and total number of brothers

```{r}
#Step 5.0.1: Counting total number of siblings
s.df %<>% mutate(hh_nsibling = hh_mtotal + hh_stotal)

#Step 5.0.2: Tabulating frequency

  ##Building table
  tab.x = data.frame(table(s.df$hh_nsibling))
  
  ##Renaming table columns
  names(tab.x) = c("Total number of siblings", "Frequency")
  
  ##Formatting table
  tab.z = kable(tab.x) %>% 
          kable_styling(full_width = FALSE, 
                        position = "left", 
                        bootstrap_options = c("striped", "hover"))

#Step 5.0.3: Generating histogram
fig.z = ggplot(data = s.df, aes(hh_nsibling)) +
        geom_histogram(binwidth = 1) +
        scale_x_continuous(breaks = seq(1, 25, 1)) +
        xlab("\nTotal Number of siblings in Household") + 
        ylab("Frequency Count\n")
```

A frequency tabulation of the total number of siblings in the household is furnished below.

`r tab.nums(name = "t04", caption = "Frequency Table, Total Number of Siblings in Household")`
```{r, echo = FALSE, results = "markdown"}
#Step 4.0.7: Printing table
tab.z
```

A histogram describing the total number of sisters in the household is furnished below.
`r fig.nums(name = "f04", caption = "Histogram, Total Number of Siblings in Household")`
```{r, echo = FALSE, results = "markdown"}
#Step 4.0.7: Printing table
fig.z
```

# Proportion of respondants living with both biological parents

Similar to previous steps, to identify this proportion we create a proxy data frame for variables asking about the relationship of the respondant to the household member if they already identified the household member as "mother" or "father", and then sum the responses for biological parents

```{r}
#Step 6.0.1: Subsetting data frame
p.df = s.df %>% select(starts_with("h1hr6"))

#Step 6.0.2: Recoding values based on codebook
p.df[p.df != 1 & p.df != 7] = 0
p.df[p.df == 1 | p.df == 7] = 1

#Step 6.0.3: Counting total number of responses indicating n'th HH member is bio parent
p.df %<>% mutate(hh_nbioparent = rowSums(p.df))

#Step 6.0.4: Binding results to main data frame
s.df %<>% mutate(hh_nbioparent = p.df$hh_nbioparent)
s.df %<>% mutate(hh_bothbiop = p.df$hh_nbioparent > 1)

#Step 6.0.5: Estimating the proportion of respondants living with 2+ bio parents
p.bioboth = round(prop.table(table(s.df$hh_bothbiop)) * 100, digits = 2)
```

In counting the number of bio parents present in the household, an anomaly was observed wherein 1 record counted 3 bio parents in the household. For the purpose of this exercise, this response is treated as valid. Thus, the proportion of respondants who live with 2 or more bio parents (e.g., "both" bio parents) is `r p.bioboth[2]`%.

# Source Code Appendix

The source code for this document is included below.
```{r comment = '', results = "markdown"}
cat(readLines(con = "estebanv_csde502_hw9.Rmd"), sep = "\n")
```

